Exploratory Data Analysis for College Scorecard Dataset

### 2017-02-22 Draft-xuezhou-sjb.Rmd
fix index and grammers

#Table of Contents
* 1.[Introduction](#Introduction)
* 2.[Goal](#Goal)
* 3.[Dataset Preparation](#dataprep)
* 4.[Objective](#objective)
* 5.[Summary of variables](#variablesummary)
* 5.1 [Individual variable summary](#5.1)
  * 5.1.1 [School](#5.1.1)
    * 5.1.1.1 [State](#5.1.1.1)
    * 5.1.1.2 [Predominant Degree](#5.1.1.2)
    * 5.1.1.3 [Control of Institution](#5.1.1.3)
  * 5.1.2 Student Body
    * 5.1.2.1 [Percentage first-generation students](#5.1.2.1)
    * 5.1.2.2 [Number of applications](#5.1.2.2)
  * 5.1.3 Cost 
    * 5.1.3.1 [Average net price for Title IV institutions(both public and private institutions)](#5.1.3.1)
    * 5.1.3.2 [Percentage of students who are financially independent](#5.1.3.2)
  * 5.1.4 Financial Aid
    * 5.1.4.1 [Percentage of undergraduates who receive a Pell Grant](#5.1.4.1)
    * 5.1.4.2 [Percent of all federal undergraduate students receiving a federal student loan](#5.1.4.2l)
    * 5.1.4.3 [The median debt for students who have completed](#5.1.4.3)
    * 5.1.4.4 [The median debt for students who have not completed](#5.1.4.4)
  * 5.1.5 Earning
    * 5.1.5.1 [Median family income in real 2015 dollars](#5.1.5.1)
    * 5.1.5.2 [Earning after graduation](#5.1.5.2)
  * 5.1.6 Repayment
    * 5.1.6.1 [3-year Repayment rate](#5.1.6.1)
    * 5.1.6.2 [3-year Default rate](#5.1.6.2)
* 6.Relationship between variables
 * 6.1 [The relationship between NPT4_COMBINE and State](#6.1)
 * 6.2 [The relationship between Debt and Cost](#6.2)
 * 6.3 [The relationship between Percentage of Pell Grant and college type(CONTROL)](#6.3)
 * 6.4 [The relationship between Percentage first-generation students and States](#6.4)
 * 6.5 [The relationship between family income and percentage of Pell Grant](#6.5)
 * 6.6 [The relationship between DEP_STAT_PCT_IND and Control](#6.6)
 * 6.7 [The relationship between PCTLOAN and Control](#6.7)
 * 6.8 [cost and median debts with earning](#6.8)
 * 6.9 [default with earning](#6.9)
 * 6.10 [degree type vs. earning](#6.10)
 * 6.11 [predominant degree, control  and earning after 6 year of entry](#6.11)
 * 6.12 [predominant degree + control (count)](#6.12)
 * 6.13 [control + earning](#6.13)
* 7. [Findings and Conclusion](#7)

#1. Introduction<a id="Introduction"></a>
The College Scorecard project was initially designed to increase the transparency of the college data, working as a navigator for students to make informed decisions about their choice for higher education. The information gathered in the dataset also gives students and families more power to differentiate postsecondary institutions and get an idea about how well colleges are preparing their students to be successful. The dataset was created to address the lack of information about college costs, earnings, federal student aid, admission rate, etc. The data provides insights into the performance of institutions that receive federal aid dollars, and the outcomes of the students of those institutions. It also gives extra support for students and families to weigh the tradeoffs on choosing the right college according to their own needs and educational goals.

The data was collected by the United States Department of Education, which corresponds to all undergraduate degree-granting institutions of higher education from 1996 through 2016. The dataset is pretty rich and covers a wide range of topics. It has 1743 associated columns. Each corresponds to one named data variable. All data variables are divided into 10 categories ranging from Root category to Earnings category. The data type contains integer, string and float. Data elements with an IPEDS value in the Source field are derived directly from the Integrated Postsecondary Education Data System (IPEDS). Further detailed description of the dataset can be found in the CollegeScorecardDataDictionary.xlsx excel file.

@@

The goal of our study is to understand what factors are predictive of students' post-graduation earnings through individual variable analysis and relationship detection. To this end, we first explored a set of variables that were carefully selected to represent a school's profile, getting to know the distribution of each variable. Then we further analyzed how these variables are related, particually the relationship between earnings and other variables.

#3. Dataset Preparation<a id="dataprep"></a>
Our analysis focuses on the data of year 2012 -2013 of 23 variables. We chose the median income 6 years after entry as our predictor variable, and eliminated many fields with a bunch of missing values as well as some fields not pertaining enougn to post-graduation income. We then fosued primarily on a set of 23 variables, including age, acedemic costs, family income, completetion rates and repayment rate.

@@

The full dataset contains 1743 variables. 36 variables were intially selected to be analyze.
```{r, message=FALSE,warning=FALSE}

@@

Considering the variable NPT4_PUB and NPT4_PRIV are mutually exclusive both indicate net price. we integrated the information of the two coloumns into a single column, named as NPT4_COMBINE. The new column records the average net price for both private and public TITLE IV institutions.
```{r, message=FALSE,warning=FALSE}

@@

In order to choose the variables that have enough valid data related to the topic, we only considered variables with percent missing value less than 25%. The new data set was named as data_clean. 
```{r, message=FALSE,warning=FALSE}

@@

The data_clean will be the clean dataset we used in the further variable summary and comparison.For numerical variables, we calculate the basic summary statistics such as median, min, max, mode and quantile. We also use visualizations such as historgams and boxplots to understand the distribution of the values. For categorical variables, we use frequency table to understand the frequencies of each category.

The table below lists the variables that will be used for the further analysis: 13 of them are continuous variables while 5 of them are categorical variables or Unique ID. After initial data cleaning process, the new data set now contains 18 variables and 7793 observations.

@@

we will summarize the data set using functions provided in dplyr and ggplot funtions in the variables summary section.

@@

Our objective at this stage is to conduct an exploration analysis on the CollegeScore dataset, investigating the individual variable's summary statistics and the relationship between these variabls, so as to better prepare for further analaysis on the factors affecting students' post college earnings with some advanced data mining techniques. The summary statistics, dplyr function, ggplot functions as well as some visualization techniques will be mainly used in this project.

#5. Summary of variables<a id="5"></a>
##5.1 Individual variable summary<a id="5.1"></a>

###5.1.1 School<a id="5.1.1"></a> 
####5.1.1.1. State<a id="5.1.1.1"></a>

State variable has 59 levels, which include 55 states in the United States and 4 districts outside mainland US. To better understand how many universities each state has, we created a frequency table as shown below, and also a bar chart to visuzalize the top 15 states having the most universities. Some basic summary statistics are also displayed.

@@

####5.1.1.3. Control of Institution<a id="5.1.1.3"></a> 
In addition to state and predominant degree, control of institution is another categorical variable in the dataset. The label of Control is to indicate whether or not an institution is privately owned and whether or not it is a for-profit organization. There are three levels of this variable: Public, Private Nonprofit, and Private-profit. Public school is assumed to be non-profit so it is not further divided into two subgroups. Below we show the percentage summary of each category and a bar chart is plotted to facilitate understanding.

@@

```
From the output, we can tell that the highest value existed in  La Belle Beauty Academy with the value of 95.25%. The lowest value is 6.87%, which existed in Wheaton College. From the mean value of 48.08%, we can tell that about for federally aided students in most colleges, around half (50.8%) of the students' parents did not complete the college. The standard deviation value of 12.2% indicated a relatively high difference of this element among different colleges.

@@

The variable APPL_SCH_PCT_GE2 is to describe the percentage of students who list two and more than two institutions when he applied for financial aid when submitting application for university. This information is only available to students who submitted their applications for federal aid and does not cover the whole student body. Nevertheless, this variable still can provide some useful insights about how many universities a student usually apply for.

@@

The summary statistics show that the median percentage of those students submitting two or more than two application is 41.67% across all universities The institution having the least percentage only has 4% of students applying for more than one university, while the school with the highest percent has as high as 90% of students applying for more than one university. The boxplot compares the percent of student applying for only one university with the percent of students applying for more than one. We can clearly see that a larger portion of students only listed one institution when submitting their application for financial aid.

###5.1.3 Cost<a id="5.1.3"></a>
####5.1.3.1 Average net price for Title IV institutions(both public and private institutions)<a id="5.1.3.1"></a>

With the followin codes, we extracted the IV institutions with the highest and lowest Average net price. Then we observed this varibles's basic summary statistics: min, max, mean, median and standard deviation.

@@

From the summary results, we learned that the maximum value of the variable is $87,414, which is for Aerosim Flight Academy located in FL. while the minimum value is about $-19,360 - the value for Berea College located in KY. Considering the elements are collected from Title IV institutions, we reasoned that the negative values may be due to the fact that some institution's financial supoort can cover the students' all the costs, even overbalancing the cost amount. There are not much differerence between the mean value and the median value, which is around $15,000. We can conclude that the average net price costs for most institutions are falling into  this range. The standard deviation value is 7604.884 which is relatively large, indicating a great spread in the data.

@@

From the histogram, we notice that the average net cost for IV institutions typically is $16000-$18000. The values between $16000 -$20000 are quite frequent. There were an unsually lagrge number of costs. The institutions with cost of more than 50000 were atypical. Due to the atypical large value, the histogram is abviously right skewed. Without those extreme values, the histogram would be reasonably symmetric.

@@

From the summary results, we learned that the maximum value of the variable is 98.56%(Capella University) while the minimum value is about 0.73%(University of Notre Dame). From the mean value, we can tell that  in most colleges, around 56% of the students are financially independent. The standard deviation value 23.3% indicates a relatively high difference among different colleges, which can also be concluded from the difference between the maximum value and the minimum value. Below is the code to generate a visual representation in the histogram

@@

From the histogram, we notice that the values between 0.7 to 0.8 are quite frequent. There was no obvious extreme value in the sample, thus the histogram is normally distributed.

@@

From the output, we can tell that the range is between 0 to 100%.	No students in Thunderbird School of Global Management has received the Pell grant while all of the students in American Conservatory Theater received the Pell Grant. Among the colleges in the sample, half of the colleges colleges have the rate as high as 52.7% (median value of 52.7%), which means more than half of undergraduare students are able to receive the Pell Grant to cover their costs. The standard deviation value is 22.57%,  indicating a relatively high difference of this element among different colleges, which can also be noticed from the largest range (100%). Below is the code to generate a visual representation in the histogram.

@@

This element, as reported in the IPEDs Student Finaicial Aid component, shows the share of undergraduate students who received federal loans in a given year. It can provide important context to figures related to debt, repayment and non-repayment.  
We are using the code below to summarize this variable and extract the institutions with highest and lowest value.

@@

From the output, we can tell that the range is between 0 to 100%.	As we can recall from the previous variable(PCTPELL), Gadsden State Community College has no students receiving federal loans in the given year. Among the colleges in the sample, half of colleges have this rate as high as 60.08% , which means around 60% of undergraduate students receive the fedaral student loans. The standard deviation value is 29.26%,  indicating a relatively high difference of this element among different colleges, which can also be noticed from the largest range (100%). Below is the code to generate a visual representation in the histogram.

@@

From the histogram, it's obvious that there are two peaks in the graph. The rate between 0-10% and the rates between 60% to 80% have similarly frequency compared to the other ranges.

@@

This variable recorded debt for students who have completed the education. We are using the code below to summarize this variable. 
```{R, message=FALSE, warinig = FALSE}

@@

We use summary statistics, standard deviation and histogram to understand the distribution of this variable. The summary statistics show that students in more than 50% of institution have a median debt of more than $5500, which is not very unaffordable. The median debt for attending students can be as much as $27620 for some school and as low as $600 for others, which indicates that it may have a large variance. So we also calculate standard deviation to measure the dispersion of the distribution. The standard deviation is $2415.623, which means about 95% of institution have  median debt ranging from $1167.754  to $10830.246. The histogram shows that the distribution of the median roughly follows a standard distribution with a sharp peak and slightly heavy tail.

@@

From the output, we can tell that the highest value is $155,606.6 for Jewish Theological Seminary of America, while the lowest value is $0 for Associated Technical College-Los Angeles. From the median value of $19,262.26, we can tell that among the colleges in the sample, the students in half of them had the family income below $19,262.26	. The standard deviation value is $18,824.31,  indicating a relatively high difference of this element among different colleges. Below is the code to generate a visual representation in the histogram

@@

From the histogram, we are able to conclude that in most colleges, students' family income fall into the range of $12,000 - $18,000. Due to some atypical large values, the histogram is obviously skewed to right. The values larger than $18,000 are tend to be extreme and the cases larger than $75000 are rare, which means that rare colleges have the students with family income more than $75000.

@@

We selected two variables MD_EARN_WNE_P6, MD_EARN_WNE_P10 in the earning group to study students' post-graduation earnings. MD_EARN_WNE_P6 is the median earning for students after six years of their entry, which corresponds to two years after graduation for a four-year program. Likewise, MD_EARN_WNE_P10 is the median earning ten years after their entry, which usually corresponds to four years after graduation These two metrics give us an idea of how much money students are making after their graduation, in both short and medium terms.

@@

The six-year after-entry median earnings of 50% institutions is below the $25,300, while ten-year-after-entry median earning is higher by about $5,500. The histogram gives us a more clear idea of the earning distribution. Most institutions with students having a median earnings between $10,000 to $50,000 six years after entry. The distribution of ten-year-after-entry falls at the right side of the six-year distribution, which means that the earning increases as students' experience increases. Besides, the ten-year post-graduation earning has a wider spread than that of the six-years. This indicates that there is a greater gap between who are earning more and who are earning less, which can be confirmed by an increase in the standard deviation.

@@

####5.1.6.1 3-year Repayment rate<a id="5.1.6.1"></a>
The variable RPY_3YR_RT_SUPP measures the percentage of students who entered repayment from October 1, 2010 to September 30, 2011, and made progress in the following three years. Repayment rates are generally considered more sensitive than default rates(we will study later), which measures only the worst-case scenario for repayment outcomes.

@@

#6. Relationship between variables<a id="6"></a>

##6.1. The relationship between average net cost for Title IV institutions and State<a id="6.1"></a>

To analyze the relationship between the average net cost for Title IV institutions and State, we grouped the data into the state level to take a further look. We would like to investigate the distribution of average net price in each state.

@@

Based on the output, we can see that the highest median value for the avreage net price exists in DC while the lowest median value is in AS(American Samoa). Moreover, we found that the top 5 states are DC, VT, MA, NH, RI, which are all the states have  well-known educational institutions.

@@

The results show that the universities located in ND, SD, VT, MN and MT have the lowest percentage first-generation students, which indicates that the students in these five states would have a better family education background. The five states with the highest share of first-generation students are LA, NJ, MP, MV, MH.

@@

##6.6. The relationship between DEP_STAT_PCT_IND and Control<a id="6.6"></a>

@@

##6.7. The relationship between PCTLOAN and Control.<a id="6.7"></a>

@@

From the result,  private institutions have a much larger median value than the public institutions. With a much higher costs burden, it makes sense that the low-income students in private institutions are much more likely to seek financial aid than the public institution students. 
From the stacked histogram, we can tell that for all  three college types, the most frequent range appears between 0.6 and 0.8, which is consistent with the findings we observe in the individual variable summary section.

##6.8 Median debts with earning<a id="6.8"></a> 
Next, we analyze the relationship between the amount the debt and post-graduation earnings. There are two variables in the dataset to summarize the median amount of debt for student body, one for students who already graduated and one for students who have not completed their study. We decided to use debt for students who have not completed their study in this problem, since it makes more sense for the purpose of prediction.

@@

##6.9 default with earning<a id="6.9"></a>

For one thing, understanding the relationship between the default rate of loan repayment and post-graduation earnings can help federal financial committee to improve their decision about load granting. If the default rate decreases as the earnings increase, the committee may prefer to grant more financial aid to institutions whose students have higher earnings. For another, the default rate of repayment for a university can also help predict the post-graduation earnings for its students.

@@

A scatter plot is created to visualize the relationship between these two variables. From the chart, we see that earnings seems to show a decreasing trend as the default rate increases, which shows a negative relationship between earnings and default rate. Then we calculate the correlation coefficient between these two variables. The resulting r = -0.33 shows that the correlation between earnings and default rate is negative but the relationship is not very strong, which may be due to the influences of some outliers. To conclude, the negative relationship between earnings and default rate of loan repayment needs extra data to confirm.

##6.10 degree type vs. earning<a id="6.10"></a>

The higher education one receives, the higher the earning is after graduation. This is an assumption we are making when investing higher education. But is it really the case. In this question, we would like to study if it's worthwhile to invest in higher degree.

@@

From the boxplot below, we see that the higher the education, the higher the median earning 6 years after entry. Especially, a graduate-degree gives a hugh rise to one's earning.

##6.11 predominant degree, control  and earning after 6 year of entry<a id="6.11"></a>

The following plot shows two factor variables (predominant degree award variable and control variable) associated with one numeric variable (MD_EARN_WNE_P6: Median earnings of students working and not enrolled 6 years after entry).

@@

Mean value of the earnings is more representative for the two categories, predominant degree and control According to the plot, we realized that private for-profit colleges whose predominant degree awarded is graduate degree has the lowest average earning with the minimum average value at $3100, while Public college whose predominant degree awarded is graduate degree has the highest average earnings with the maximum average value at $45,850.

##6.12 predominant degree + control (count)<a id="6.12"></a>

@@

For the R code, our team first set the aesthetic mapping on Control_factor as the horizontal axis. Then, we use the function geom_bar() to fill different control types using the counting values of PREDDEG, which creates bars based on a grouped frequency distribution. After that, we further use the function xtabs() to calculate the exact number for each predominant degree awarded by different types of colleges.

@@

#6.13 control + earning<a id="6.13"></a>

@@

All the three types of school have far out values above, which can be explored further. Obvious differences contain in these three boxplots. The boxplot of public college is comparatively short, while the boxplot of private nonprofit is comparatively tall. The private for-profit college has the lowest median earning.

>>>>

#Table of Contents
* 1.[Introduction](#Introduction)
* 2.[Goal](#Goal)
* 3.[Dataset Preparation](#dataprep)
* 4.[Objective](#objective)
* 5.[Summary of Variables](#variablesummary)
* 5.1 [Individual Variable Summary](#5.1)
  * 5.1.1 [School](#5.1.1)
    * 5.1.1.1 [State](#5.1.1.1)
    * 5.1.1.2 [Predominant Degree](#5.1.1.2)
    * 5.1.1.3 [Control of Institution](#5.1.1.3)
  * 5.1.2 Student Body
    * 5.1.2.1 [Percentage First-generation Students](#5.1.2.1)
    * 5.1.2.2 [Number of Applications](#5.1.2.2)
  * 5.1.3 Cost 
    * 5.1.3.1 [Average Net Price for Title IV Institutions (Both Public and Private Institutions)](#5.1.3.1)
    * 5.1.3.2 [Percentage of Students Who are Financially Independent](#5.1.3.2)
  * 5.1.4 Financial Aid
    * 5.1.4.1 [Percentage of Undergraduates Who Receive a Pell Grant](#5.1.4.1)
    * 5.1.4.2 [Percent of all Federal Undergraduate Students Receiving a Federal Student Loan](#5.1.4.2l)
    * 5.1.4.3 [The Median Debt for Students Who Have Completed](#5.1.4.3)
    * 5.1.4.4 [The Median Debt for Students Who Have not Completed](#5.1.4.4)
  * 5.1.5 Earning
    * 5.1.5.1 [Median Family Income in real 2015 Dollars](#5.1.5.1)
    * 5.1.5.2 [Earning after graduation](#5.1.5.2)
  * 5.1.6 Repayment
    * 5.1.6.1 [3-year Repayment rate](#5.1.6.1)
    * 5.1.6.2 [3-year Default rate](#5.1.6.2)
* 6.Relationship between variables
 * 6.1 [The Relationship between NPT4_COMBINE and State](#6.1)
 * 6.2 [The Relationship between Debt and Cost](#6.2)
 * 6.3 [The Relationship between Percentage of Pell Grant and College Type (CONTROL)](#6.3)
 * 6.4 [The Relationship between Percentage First-generation Students and States](#6.4)
 * 6.5 [The Relationship between Median Family Income and Percentage of Pell Grant](#6.5)
 * 6.6 [The Relationship between the Percentage of Financially-independent Students and College Type (Control)](#6.6)
 * 6.7 [The Relationship between Percentage of Students Who Receive the Federal Loans and College Type (Control).](#6.7)
 * 6.8 [The Relationsip between Median Debts and Post-Graduation Earnings](#6.8)
 * 6.9 [The Relashionship between the Default Rate of Loan Repayment and Post-graduation Earnings ](#6.9)
 * 6.10 [The Relashionship between Earning 6 Years from Entry and Degree Type](#6.10)
 * 6.11 [The Relashionship among Predominant Degree, College Type(Control) and Earning after 6 year of Entry](#6.11)
 * 6.12 [The Relashionship between Predominant Degree and College Type (Control)](#6.12)
 * 6.13 [The Relashionship between College Type (Control) and Median earnings of Students Working and Not Enrolled 6 Years after Entry](#6.13)
* 7. [Findings and Conclusion](#7)

#1. Introduction<a id="Introduction"></a>
The College Scorecard project was initially designed to increase the transparency of the college data, working as a navigator for students to make informed decisions about their choice for higher education. The information gathered in the dataset also gives students and families more power to differentiate postsecondary institutions and get an idea about how well colleges are preparing their students to be successful. The dataset was created to address the lack of information about college costs, earnings, federal student aid, admission rate, etc. The data provides insights into the performance of institutions that receive federal aid dollars, and the outcomes of the students of those institutions. It also gives extra support for students and families to weigh the tradeoffs on choosing the right colleges according to their own needs and educational goals.

The data was collected by the United States Department of Education, which corresponds to all undergraduate degree-granting institutions of higher education from 1996 through 2016. The dataset is pretty rich and covering a wide range of topics. It has 1743 associated columns, each corresponds to one named data variable. All data variables are divided into 10 categories ranging from Root category to Earnings category. The data type contains integer, string and float. Data elements with an IPEDS value in the Source field are derived directly from the Integrated Postsecondary Education Data System (IPEDS). Further detailed description of the dataset can be found in the CollegeScorecardDataDictionary.xlsx excel file.

@@

The goal of our study is to understand what factors are predictive of students' post-graduation earnings through individual variable analysis and relationship detection. To this end, we first explored a set of variables that were carefully selected to represent a school's profile and students demographic information, getting to know the distribution of each variable. Then we further analyzed how these variables are related, particually the relationship between earnings and other variables.

#3. Dataset Preparation<a id="dataprep"></a>
Our analysis focuses on the data of year 2012 -2013 of 38 variables. We chose the median income 6 years after entry as our predictor variable, and eliminated many fields with a bunch of missing values as well as some fields not pertaining enougn to post-graduation income. We then fosued primarily on a set of 23 variables, including age, acedemic costs, family income, completetion rates and repayment rate.

@@

The full dataset contains 1743 variables, as we metioned before, 38 variables were intially selected to be analyze.
```{r, message=FALSE,warning=FALSE}

@@

Considering the variable NPT4_PUB and NPT4_PRIV are mutually exclusive, which both represent average net price, we integrated the information of the two coloumns into a single column, named as NPT4_COMBINE. The new column records the average net price for both private and public TITLE IV institutions.
```{r, message=FALSE,warning=FALSE}

@@

In order to choose the variables that have enough valid observations related to the topic, we only considered variables with the percent missing value less than 25%. The new data was named as data_clean. 
```{r, message=FALSE,warning=FALSE}

@@

For numerical variables, we calculated the basic summary statistics such as median, min, max, standard deviation and mean. We also used visualizations such as historgams and boxplots to understand the distribution of the values. For categorical variables, we used frequency table to understand the frequencies of each category.

The table below listed the variables that will be used for the further analysis: 13 of them are continuous variables while 5 of them are categorical variables or Unique ID. After initial data cleaning process, the new data set now contains 18 variables and 7793 observations.

@@

we summarized the data set using functions provided in dplyr and ggplot in the variables summary section.

@@

Our objective at this stage is to conduct an exploratory data analysis on the CollegeScore dataset, investigating the individual variables' summary statistics and the relationship between these variabls, so as to better prepare for further analaysis on the factors affecting students' post college earnings with some advanced data mining techniques. The summary statistics, dplyr function, ggplot functions as well as some visualization techniques were mainly used in this project.

#5. Summary of variables<a id="5"></a>
##5.1 Individual variable summary<a id="5.1"></a> 
###5.1.1 School<a id="5.1.1"></a> 
####5.1.1.1. State<a id="5.1.1.1"></a>

The State variable has 59 levels, which include 55 states in the United States and 4 districts outside mainland US. To better understand how many universities each state has, we created a frequency table as shown below, and also a bar chart to visuzalize the top 15 states having the most universities. Some basic summary statistics are also displayed.

@@

####5.1.1.3. Control of Institutions <a id="5.1.1.3"></a> 
In addition to state and predominant degree, control of institution is another categorical variable in the dataset. The label of Control is to indicate whether or not an institution is privately owned and whether or not it is a for-profit organization. There are three levels of this variable: Public, Private Nonprofit, and Private-profit. Public school is assumed to be non-profit so it is not further divided into two subgroups. Below we show the percentage summary of each category and a bar chart is plotted to facilitate understanding.

@@

```
From the output, we can tell that the highest value existed in  La Belle Beauty Academy with the value of 95.25%. The lowest value is 6.87%, which existed in Wheaton College. From the mean value of 48.08%, we can tell that for federally aided students in most colleges, around half (50.8%) of the students' parents did not complete the college. The standard deviation value of 12.2% indicated a relatively high difference of this element among different colleges.

@@

The variable APPL_SCH_PCT_GE2 is to describe the percentage of students who list two and more than two institutions when he applied for financial aid when submitting application for university. This information is only available to the students who submitted their applications for federal aid and does not cover the whole student body. Nevertheless, this variable still can provide some useful insights about how many universities a student usually apply for.

@@

The summary statistics show that the median percentage of those students submitting two or more than two application is 41.67% across all universities. The institution having the least percentage only has 4% of students applying for more than one university, while the school with the highest percent has as high as 90% of students applying for more than one university. The boxplot compares the percent of students applying for only one university with the percent of students applying for more than one. We can clearly see that a larger portion of students only listed one institution when submitting their application for financial aid.

###5.1.3 Cost<a id="5.1.3"></a>
####5.1.3.1 Average Net Price for Title IV Institutions(both public and private institutions)<a id="5.1.3.1"></a>

With the following codes, we extracted the IV institutions with the highest and lowest average net price. Then we observed this varibles's basic summary statistics: min, max, mean, median and standard deviation.

@@

From the summary results, we learned that the maximum value of the variable is $87,414, which is for Aerosim Flight Academy located in FL. while the minimum value is about $-19,360 - the value for Berea College located in KY. Considering the elements are collected from Title IV institutions, we reasoned that the negative values may be due to the fact that some institution's financial supoort can cover the students' all the costs, even overbalancing the cost amount. There are not much differerences between the mean value and the median value, which is around $15,000. We can conclude that the average net price costs for most institutions are falling into  this range. The standard deviation value is $7604.884 which is relatively large, indicating a great spread in the data.

@@

From the histogram, we notice that the average net cost for IV institutions typically is $16000-$18000. The values between $16000 -$20000 are quite frequent. There were an unsually lagrge number of costs. The institutions with cost of more than $50000 were atypical. Due to the atypical large value, the histogram is abviously right skewed. Without those extreme values, the histogram would be reasonably symmetric.

@@

From the summary results, we learned that the maximum value of the variable is 98.56%(Capella University) while the minimum value is about 0.73% (University of Notre Dame). From the mean value, we can tell that in most colleges, around 56% of the students are financially independent. The standard deviation value 23.3% indicates a relatively high difference among different colleges, which can also be concluded from the difference between the maximum value and the minimum value. Below is the code to generate a visual representation in the histogram

@@

From the histogram, we notice that the values between 0.7 to 0.8 are quite frequent. There is no obvious extreme value in the sample, thus the histogram is normally distributed.

@@

From the output, we can tell that the range is between 0 to 100%.	No students in Thunderbird School of Global Management has received the Pell grant while all of the students in American Conservatory Theater received the Pell Grant. Among the colleges in the sample, half of the colleges have the rate as high as 52.7% (median value of 52.7%), which means more than half of undergraduare students are able to receive the Pell Grant to cover their costs. The standard deviation value is 22.57%,  indicating a relatively high difference of this element among different colleges, which can also be noticed from the largest range (100%). Below is the code to generate a visual representation in the histogram.

@@

This element, as reported in the IPEDs Student Finaicial Aid component, shows the share of undergraduate students who received federal loans in a given year. It can provide important context to figures related to debt, repayment and non-repayment.We are using the code below to summarize this variable and extract the institutions with highest and lowest value.

@@

From the output, we can tell that the range is between 0 to 100%.	As we can recall from the previous variable (PCTPELL), Gadsden State Community College has no students receiving federal loans in the given year. Among the colleges in the sample, half of colleges have this rate as high as 60.08% , which means around 60% of undergraduate students received the fedaral student loans in the given year. The standard deviation value is 29.26%,  indicating a relatively high difference of this element among different colleges, which can also be noticed from the largest range (100%). Below is the code to generate a visual representation in the histogram.

@@

From the histogram, it's obvious that there are two peaks in the graph. The rate between 0-10% and the rates between 60% to 80% have similar frequency compared to the other ranges.

@@

This variable recorded debt for students who have completed the education. We are using the code below to summarize this variable and extracted the institutions with largest and smallest value.

@@

We use summary statistics, standard deviation and histogram to understand the distribution of this variable. The summary statistics show that students in more than 50% of institutions have a median debt of more than $5500, which is not very unaffordable. The median debt for attending students could be as much as $27620 for some schools and as low as $600 for others, indicating that it may have a large variance. So we also calculate standard deviation to measure the dispersion of the distribution. The standard deviation is $2415.623, which means about 95% of institution have  median debt ranging from $1167.754  to $10830.246. The histogram shows that the distribution of the median roughly follows a standard distribution with a sharp peak and slightly heavy tail.

@@

From the output, we can tell that the highest value is $155,606.6 for Jewish Theological Seminary of America, while the lowest value is $0 for Associated Technical College-Los Angeles. From the median value of $19,262.26, we can tell that among the colleges in the sample, the students in half of them had the family income below $19,262.26. The standard deviation value is $18,824.31, indicating a relatively high difference of this element among different colleges. Below is the code to generate a visual representation in the histogram

@@

From the histogram, we are able to conclude that in most colleges, students' family income fall into the range of $12,000 - $18,000. Due to some atypical large values, the histogram is obviously skewed to right. The values larger than $18,000 are tending to be extreme and the cases larger than $75000 are rare, which means that rare colleges have the students with family income more than $75000.

@@

MD_EARN_WNE_P6 and MD_EARN_WNE_P10 are two of the earning varibles we selected to study students' post-graduation earnings. MD_EARN_WNE_P6 is the median earning for students after six years of their entry, which corresponds to two years after graduation for a four-year program. Likewise, MD_EARN_WNE_P10 is the median earning ten years after their entry, which usually corresponds to four years after graduation. These two metrics give us an idea of how much money students are making after their graduation, in both short and medium terms.

@@

The six-year after-entry median earnings of 50% institutions is below the $25,300, while ten-year-after-entry median earning is higher by about $5,500. The histogram gives us a more clear idea of the earning distribution. Most institutions with students having a median earnings between $10,000 to $50,000 six years after entry. The distribution of ten-year-after-entry falls at the right side of the six-year distribution, which means that the earning increases as students' experience increases. Besides, the ten-year post-graduation earning has a wider spread than that of the six-years. This indicates that there is a greater gap of the earning amount among the different institutions, which can be confirmed by an increase in the standard deviation.

@@

####5.1.6.1 3-year Repayment rate<a id="5.1.6.1"></a>
The variable RPY_3YR_RT_SUPP measures the percentage of students who entered repayment from October 1, 2010 to September 30, 2011 and made progress in the following three years. Repayment rates are generally considered more sensitive than default rates(we will study later), which measures only the worst-case scenario for repayment outcomes.

@@

#6. Relationship between variables<a id="6"></a>
##6.1. The relationship between average net cost for Title IV institutions and State<a id="6.1"></a>

To analyze the relationship between the average net cost for Title IV institutions and State, we grouped the data into the state level to take a further look. We would like to investigate the distribution of institutions ' average net price in each state.

@@

Based on the output, we can see that the highest median value for the avreage net price exists in DC while the lowest median value is in AS(American Samoa). Moreover, we found that the top 5 states are DC, VT, MA, NH, RI, which are the states with certain well-known educational institutions.

@@

The results show that the universities located in ND, SD, VT, MN and MT have the lowest percentage first-generation students, which indicates that the students in these five states would have a better family education background. The five states with the highest share of first-generation students are LA, NJ, MP, MV, and MH.

@@

##6.6. The Relationship between the Percentage of Financially-independent Students and Control<a id="6.6"></a>

@@

##6.7. The Relationship between Percentage of Students Who Receive the Federal Loans and Control.<a id="6.7"></a>

@@

From the result,  private institutions have a much larger median value than the public institutions. With a much higher costs burden, it makes sense that the low-income students in private institutions are much more likely to seek financial aids than the students in public institutions.

From the stacked histogram, we can tell that for all  three college types, the most frequent range appears between 0.6 and 0.8, which is consistent with the findings we observe in the individual variable summary section.

##6.8 The Relationsip between Median Debts and Post-Graduation Earnings<a id="6.8"></a> 
Next, we analyze the relationship between the amount of the debt and post-graduation earnings. There are two variables in the dataset to summarize the median amount of debt for student body, one for students who already graduated and one for students who have not completed their study. We decided to use debt for students who have not completed their study in this problem, since it makes more sense for the purpose of prediction.

@@

##6.9 The Relashionship between the Default Rate of Loan Repayment and Post-graduation Earnings<a id="6.9"></a> 
For one thing, understanding the relationship between the default rate of loan repayment and post-graduation earnings can help federal financial committee to improve their decisions about load granting. If the default rate decreases as the earnings increases, the committee may prefer to grant more financial aid to institutions whose students have higher earnings. For another, the default rate of repayment for a university can also help predict the post-graduation earnings for its students.

@@

A scatter plot is created to visualize the relationship between these two variables. From the chart, we see that earnings seems to show a decreasing trend as the default rate increases, which shows a negative relationship between earnings and default rate. Then we calculate the correlation coefficient between these two variables. The resulting r = -0.33 shows that the correlation between earnings and default rate is negative but the relationship is not very strong, which may be due to the influences of some outliers. To conclude, the negative relationship between earnings and default rate of loan repayment needs extra data to validate.

##6.10 The Relashionship between Earning 6 Years from Entry and Degree Type<a id="6.10"></a>

The higher education one receives, the higher the earning is after graduation. This is a common assumption we are making when investing higher education, and this is it really the case. In this question, we would like to study if it's worthwhile to invest in higher degree.

@@

From the boxplot above, we see that the higher the education, the higher the median earning 6 years after entry. Especially, a graduate-degree gives a hugh rise to one's earning.

##6.11 The Relashionship among Predominant Degree, Control and Earning after 6 year of Entry<a id="6.11"></a>

The following plot shows the relationship among two factor variables (predominant degree award variable and control variable) associated with one numeric variable (MD_EARN_WNE_P6: Median earnings of students working and not enrolled 6 years after entry).

@@

Mean value of the earnings is more representative for the two categories, predominant degree and control.  According to the plot, we realized that private for-profit colleges whose predominant degree awarded is graduate degree have the lowest average earning with the minimum average value at $3100, while Public colleges whose predominant degree awarded are graduate degree has the highest average earnings with the maximum average value at $45,850.

##6.12 The Relashionship between Predominant Degree and Control<a id="6.12"></a>

@@

For the R code, our team first set the aesthetic mapping on Control_factor as the horizontal axis. Then, we used the function geom_bar() to fill different control types using the counting values of PREDDEG, which created bars based on a grouped frequency distribution. After that, we used the function xtabs() to calculate the exact number for each predominant degree awarded by different types of colleges.

@@

#6.13 The Relashionship between Control and Median earnings of Students Working and Not Enrolled 6 Years after Entry<a id="6.13"></a>

@@

All the three types of schools have far out values above, which can be explored further. Obvious differences were contained in these three boxplots. The boxplot of public college is comparatively short, while the boxplot of private nonprofit is comparatively tall. The private for-profit college has the lowest median earning.
